Biogeography and Environmental Conditions Shape Phage and Bacteria Interaction Networks Across the Healthy Human Microbiome

### 2017-04-25 doc/manuscript.Rmd
Updated with Pats suggestions.
Updates updates updates.

# Abstract
Viruses and bacteria are critical components of the human microbiome and play important roles in health and disease. Most previous work has relied on studying bacteria and phages in isolation, reducing them to two separate communities. Such approaches are not able to capture how these communities interact to maintain stability and share information (i.e. horizontal gene transfer). We developed and implemented a network-based analytical approach to provide an initial understanding of phage-bacteria network diversity throughout the human body. We accomplished this by building a machine learning algorithm to predict which phages infected which bacteria in a given microbiome. This algorithm was applied to paired viral and bacterial metagenomic sequence sets from three previously published human cohorts. We organized the predicted interactions into networks that allowed us to evaluate the diversity of phage-bacteria connectedness across the human body. We found that gut and skin network structures were person-specific, and were not conserved among cohabitating family members. High-fat diets and obesity were associated with less connected networks. There were significant differences in network structure between skin sites, with those exposed to the external environment being less connected and more prone to instability. Together this study characterized the diversity of virome-microbiome networks across the human body and illustrated how environmental factors may influence phage-bacteria interactive dynamics. This work provides a baseline for future studies to investigate the role of ecological networks in conditions such as disease states.

@@

We investigated human microbiome-virome networks by utilizing three published microbiome datasets with paired virus and bacterial metagenomic sequence sets [@Hannigan:2015fz; @Minot:2011ez; @Reyes:2010cwa; @Turnbaugh:2009ei]. Our approach built off of previous large-scale phage-bacteria microbiome network analyses by inferring interactions using metagenomic datasets, instead of using culture-based techniques [@Flores:2013hc]. This metagenomic interaction inference model was powered beyond previous models by its inclusion of protein interaction data as well as both negative and positive interactions [@Edwards:2015iz, @Roux:2016cc]. Through this approach, we were able to go beyond contemporary isolated methods to provide a basic understanding of the community network dynamics associated with healthy human phages and bacteria. By building and utilizing a microbiome-virome network, we found that different people, body sites, and anatomical locations not only support distinct microbiome membership and diversity [@Hannigan:2015fz; @Minot:2011ez; @Reyes:2010cwa; @Turnbaugh:2009ei; @Grice:2009eea; @Findley:2013jf; @Costello:2009im, @Consortium:2012iz], but also support communities with distinct communication structures and propensities toward community instability. Through an improved understanding of the healthy state of network structures across the human body, we will empower future studies to begin investigating how these community structures change in disease and their overall impact on human health.

@@

The bacterial and viral sequences from these publications were quality filtered and assembled into contigs. We generated clusters of related bacterial and phage contigs by k-mer frequency and co-abundance using the CONCOCT algorithm **(Supplemental Figure \ref{ContigStats} - \ref{ClusterStats})**. We interpreted these clusters as operational units of related bacteria and phage genomes, which we defined as operational genomic units (OGUs). These OGUs are conceptually similar to the operational taxonomic unit (OTU) and operational protein family (OPF) definitions used for grouping highly similar 16S rRNA gene and open reading frame sequences, respectively [@Schloss:2008ht].

@@

Four phage and bacterial genomic features were used to predict infectious relationships between bacteria and phages: 1) genome nucleotide similarities, 2) gene amino acid sequence similarities, 3) Clustered Regularly Interspaced Short Palindromic Repeat (CRISPR) targeting of phages by bacterial CRISPR spacer sequences, and 4) similarity of protein families associated with experimentally identified protein-protein interactions [@Orchard:2014hq]. The resulting random forest model performed with an AUC of `r signif(genmodelper$ROC, digits = sigfig)`, a sensitivity of `r signif(genmodelper$Sens, digits = sigfig)`, and a specificity of `r signif(genmodelper$Spec, digits = sigfig)` **(Figure \ref{RocCurve} A)**. The most important predictor in the model was nucleotide similarity between genes, followed by nucleotide similarity of whole genomes **(Figure \ref{RocCurve} B)**. Protein family interactions were moderately important to the model, while CRISPRs were not informative **(Figure \ref{RocCurve} B)**. Approximately one third of the training set relationships yielded no score and were therefore unable to be assigned an interaction prediction **(Figure \ref{RocCurve} C)**.

Once our random forest model was built, we used it to classify the relationships between bacteria and phage operational genomic units in the experimental datasets. The relationships predicted within the three published datasets were used to construct one master network. The master network contained the three studies as sub-networks, which themselves each contained sub-networks for each sample **(Figure \ref{RocCurve} D)**. Metadata including study, sample ID, disease, and OGU abundance within the community were stored in the master network for downstream analysis **(Supplemental Figure \ref{NetworkDiagram})**. The master network was highly connected and contained `r prettyNum(genfigurestats[c(genfigurestats$class %in% "Total"),"Edges"], big.mark = ",")` infectious relationships among `r prettyNum(genfigurestats[c(genfigurestats$class %in% "Total"),"Vertices"], big.mark = ",")` nodes, `r prettyNum(nestats[c(nestats$cats %in% "PhageNodes"),"values"], big.mark = ",")` phages and `r prettyNum(nestats[c(nestats$cats %in% "BacteriaNodes"),"values"], big.mark = ",")` bacteria. Although the network was highly connected, not all relationships were present in all samples. As relationships were weighted by the relative abundances of their associated bacteria and phages, lowly abundant relationships could be present but not highly abundant. Like the master network, the skin network exhibited a diameter of `r prettyNum(genfigurestats[c(genfigurestats$class %in% "SkinStudy"),"Diameter"], big.mark = ",")` (measure of graph size; the greatest number of traversed vertices required between two vertices) and included `r round(100 * genfigurestats[c(genfigurestats$class %in% "SkinStudy"),"Vertices"] / genfigurestats[c(genfigurestats$class %in% "Total"),"Vertices"], digits = 1)`% and `r round(100 * genfigurestats[c(genfigurestats$class %in% "SkinStudy"),"Edges"] / genfigurestats[c(genfigurestats$class %in% "Total"),"Edges"], digits = 1)`% of the master network nodes and edges, respectively **(Figure \ref{RocCurve} E - F)**. The phages and bacteria in the gut diet and twin sample sets were more sparsely related, each contained fewer than 150 vertices, fewer than 20,000 relationships, and diameters of `r prettyNum(genfigurestats[c(genfigurestats$class %in% "DietStudy"),"Diameter"], big.mark = ",")`, which indicated more sparsely related phages and bacteria **(Figure \ref{RocCurve} E - F)**.

@@

We evaluated the diet-associated differences in gut virome-microbiome network structure by quantifying how central each sample's network was on average. We accomplished this by utilizing two common centrality metrics: degree centrality and closeness centrality. Degree centrality, the simplest centrality metric, was defined as the number of connections each phage made to bacteria, or each bacterium made to phages. We supplemented measurements of degree centrality with measurements of closeness centrality. Closeness centrality is a metric of how close each phage or bacterium is to all of the other phages and bacteria in the network. A higher closeness centrality suggests that the effects of genetic information or altered abundance would be more impactful to all other microbes in the system. A network with higher average closeness centrality also indicates an overall greater degree of connections, which suggests a greater resilience against instability. Because these values are assigned to each phage and bacterium within each network, we calculated the average connectedness and corrected for the maximum potential degree of connectedness to obtain a single value for the connectedness of each graph.

@@

## Individuality of Microbial Networks
Skin and gut community membership and diversity are highly personal, with people remaining more similar to themselves over time compared to other people [@Grice:2009ee; @Hannigan:2015fz; @Minot:2013ih]. We therefore hypothesized that this personal conservation extended to microbiome network structure. To answer this hypothesis, we calculated the degree of dissimilarity between each subject's network, based on phage and bacteria abundance as well as centrality. We quantified phage and bacteria centrality within each sample graph using the weighted eigenvector centrality metric. This metric defines central phages as those that are highly abundant and infect many distinct bacteria which themselves are abundant and infected by many other phages. Similarly, bacterial centrality was defined as those bacteria that were both abundant and connected to numerous phages that were themselves connected to many bacteria. We then calculated the similarity of community networks using the weighted eigenvector centrality of all nodes between all samples. Samples with similar network structures were interpreted as having similar capacities for maintaining stability and transmitting genetic material.

We used this network dissimilarity metric to test whether microbiome network structures were more similar within people than between people over time. We found that gut microbiome network structure clustered by person (ANOSIM p-value = `r signif(interstats[c(interstats$site %in% "DietAnosim"), "prob"], digits = sigfig)`, R = `r signif(interstats[c(interstats$site %in% "AS"), "prob"], digits = sigfig)`, **Figure \ref{intradiv} A**). Network dissimilarity within each person over the 8-10 day sampling period was less than the average dissimilarity between that person and others, although this difference was not statistically significant (p-value = `r signif(interstats[c(interstats$site %in% "Diet"), "prob"], digits = sigfig)`, **Figure \ref{intradiv} B**). The lack of statistical confidence was likely due to the small sample size of this dataset. Although there was evidence for gut network conservation among individuals, we found no evidence for conservation of gut network structures within families. The gut network structures were not more similar within families (twins and their mothers; intrafamily) compared to other families (inter-family) (p-value = `r signif(interstats[c(interstats$site %in% "Twins"), "prob"], digits = sigfig)`, **Figure \ref{intradiv} C**).

@@

We found evidence that diet was sufficient to alter gut microbiome network connectivity. Although our sample size was small, our findings suggested that high-fat diets were less connected than low-fat diets. This suggested that high-fat diets may lead to less stable communities with a decreased ability for individuals to influence one other. We supported this finding with the observation that obesity was associated with decreased network connectivity (closeness of bacterial and phage vertices). Together these findings suggest the food we eat may not only impact which microbes colonize our guts, but may also impact their interactions. Further work will be required to characterize these relationships with a larger cohort.

In addition to diet, the skin microbiome network structure varied by skin environment. Network structure differed between environmentally exposed and occluded skin sites. The sites under greater environmental fluctuation and exposure, the exposed and intermittently exposed sites, were less connected and therefore were predicted to have a higher propensity for instability. Likewise, intermittently moist sites demonstrated less connectedness than the more stable moist and sebaceous sites. Thus body sites under greater degrees of fluctuation harbored less connected, potentially less stable microbiomes.

Our results represent a step toward understanding the microbiome through interspecies relationships. Although these findings are informative, there are certainly caveats that should be noted. First, although our infection classification model is advantageous over existing models, we recognize that, like most classification models, ours is only as good as its training set. Large scale experimental screens for phage and bacteria infectious interactions will provide more robust model training and therefore improved model performance. Just as we have improved on previous modeling efforts, we expect that new and creative scoring metrics will be integrated into this model to further improve model performance. Second, although our analyses offer an informative proof of concept, this work was done retrospectively and relied on published research from as long as seven years ago. These archived datasets were limited by the technology and costs of the time, meaning the datasets were poorly powered for the statistical analysis we strive for today. Despite these limitations, we were able to present initial conclusions; however, follow-up studies will be required to validate our findings. These methods and results will be important for informing design and interpretation of future studies and for highlighting the importance of bacteria-virus interaction profiles as a characteristic of human-associated microbial communities.

@@

# Acknowledgments
We thank the members of the Schloss lab for their underlying contributions. GDH is supported in part by the University of Michigan Molecular Mechanisms of Microbial Pathogenesis Fellowship. GDH and PDS were supported in part by funding from the NIH (P30DK034933, U19AI09087, and U01AI124255).

>>>>

# Abstract
Viruses and bacteria are critical components of the human microbiome and play important roles in health and disease. Most previous work has relied on studying bacteria and phages in isolation, reducing them to two separate communities. Such approaches are not able to capture how these communities interact to maintain stability and share information (i.e. horizontal gene transfer). We developed and implemented a network-based analytical approach to provide an initial understanding of phage-bacteria network diversity throughout the human body. We accomplished this by building a machine learning algorithm to predict which phages infected which bacteria in a given microbiome. This algorithm was applied to paired viral and bacterial metagenomic sequence sets from three previously published human cohorts. We organized the predicted interactions into networks that allowed us to evaluate phage-bacteria connectedness across the human body. We found that gut and skin network structures were person-specific, and were not conserved among cohabitating family members. High-fat diets and obesity were associated with less connected networks. There were significant differences in network structure between skin sites, with those exposed to the external environment being less connected and more prone to instability. Together this study characterized the diversity of virome-microbiome networks across the human body and illustrated how environmental factors may influence phage-bacteria interactive dynamics. This work provides a baseline for future studies to investigate the role of ecological networks in conditions such as disease states.

@@

We investigated human microbiome-virome networks by utilizing three published microbiome datasets with paired virus and bacterial metagenomic sequence sets [@Hannigan:2015fz; @Minot:2011ez; @Reyes:2010cwa; @Turnbaugh:2009ei]. Our approach built off of previous large-scale phage-bacteria microbiome network analyses by inferring interactions using metagenomic datasets, instead of using culture-based techniques [@Flores:2013hc]. This metagenomic interaction inference model was powered beyond previous models by its inclusion of protein interaction data as well as both negative and positive interactions [@Edwards:2015iz, @Roux:2016cc]. Through this approach, we were able to go beyond contemporary isolated methods to provide a basic understanding of the network dynamics associated with healthy human phage and bacteria communities. By building and utilizing a microbiome-virome network, we found that different people, body sites, and anatomical locations not only support distinct microbiome membership and diversity [@Hannigan:2015fz; @Minot:2011ez; @Reyes:2010cwa; @Turnbaugh:2009ei; @Grice:2009eea; @Findley:2013jf; @Costello:2009im, @Consortium:2012iz], but also support communities with distinct communication structures and propensities toward community instability. Through an improved understanding of the healthy state of network structures across the human body, we will empower future studies to begin investigating how these community structures change in disease and their overall impact on human health.

@@

The bacterial and viral sequences from these publications were quality filtered and assembled into contigs. We generated clusters of related bacterial and phage contigs by k-mer frequency and co-abundance using the CONCOCT algorithm **(Supplemental Figure \ref{ContigStats} - \ref{ClusterStats})**. Contig lengths within clusters ranged between $10_{3}$ and $10_{5.5}$, confirming the majority of contigs did not represent whole genomes alone **(Supplemental Figure \ref{ContigStats} - \ref{ClusterStats})**. We interpreted these clusters as operational units of related bacteria and phage genomes, which we defined as operational genomic units (OGUs). These OGUs are conceptually similar to the operational taxonomic unit (OTU) and operational protein family (OPF) definitions used for grouping highly similar 16S rRNA gene and open reading frame sequences, respectively [@Schloss:2008ht].

@@

Four phage and bacterial genomic features were used to predict infectious relationships between bacteria and phages: 1) genome nucleotide similarities, 2) gene amino acid sequence similarities, 3) Clustered Regularly Interspaced Short Palindromic Repeat (CRISPR) targeting of phages by bacterial CRISPR spacer sequences, and 4) similarity of protein families associated with experimentally identified protein-protein interactions [@Orchard:2014hq]. The resulting random forest model performed with an AUC of `r signif(genmodelper$ROC, digits = sigfig)`, a sensitivity of `r signif(genmodelper$Sens, digits = sigfig)`, and a specificity of `r signif(genmodelper$Spec, digits = sigfig)` **(Figure \ref{RocCurve} A)**. The most important predictor in the model was amino acid similarity between genes (blastx), followed by nucleotide similarity of whole genomes **(Figure \ref{RocCurve} B)**. Protein family interactions were moderately important to the model, and CRISPRs were not informative **(Figure \ref{RocCurve} B)**. Approximately one third of the training set relationships yielded no score and were therefore unable to be assigned an interaction prediction **(Figure \ref{RocCurve} C)**.

Once our random forest model was built, we used it to classify the relationships between bacteria and phage operational genomic units in the experimental datasets. The relationships predicted within the three published datasets were used to construct one master network. The master network contained the three studies as sub-networks, which themselves each contained sub-networks for each sample **(Figure \ref{RocCurve} D)**. Metadata including study, sample ID, disease, and OGU abundance within the community were stored in the master network for downstream analysis **(Supplemental Figure \ref{NetworkDiagram})**. The master network was highly connected and contained `r prettyNum(genfigurestats[c(genfigurestats$class %in% "Total"),"Edges"], big.mark = ",")` infectious relationships among `r prettyNum(genfigurestats[c(genfigurestats$class %in% "Total"),"Vertices"], big.mark = ",")` nodes, `r prettyNum(nestats[c(nestats$cats %in% "PhageNodes"),"values"], big.mark = ",")` phages and `r prettyNum(nestats[c(nestats$cats %in% "BacteriaNodes"),"values"], big.mark = ",")` bacteria. Although the network was highly connected, not all relationships were present in all samples. As relationships were weighted by the relative abundances of their associated bacteria and phages, lowly abundant relationships could be present but not highly abundant. Like the master network, the skin network exhibited a diameter of `r prettyNum(genfigurestats[c(genfigurestats$class %in% "SkinStudy"),"Diameter"], big.mark = ",")` (measure of graph size; the greatest number of traversed vertices required between two vertices) and included `r round(100 * genfigurestats[c(genfigurestats$class %in% "SkinStudy"),"Vertices"] / genfigurestats[c(genfigurestats$class %in% "Total"),"Vertices"], digits = 1)`% and `r round(100 * genfigurestats[c(genfigurestats$class %in% "SkinStudy"),"Edges"] / genfigurestats[c(genfigurestats$class %in% "Total"),"Edges"], digits = 1)`% of the master network nodes and edges, respectively **(Figure \ref{RocCurve} E - F)**. The phages and bacteria in the gut diet and twin sample sets were more sparsely related, with each containing fewer than 150 vertices, fewer than 20,000 relationships, and diameters of `r prettyNum(genfigurestats[c(genfigurestats$class %in% "DietStudy"),"Diameter"], big.mark = ",")`, which indicated more sparsely related phages and bacteria **(Figure \ref{RocCurve} E - F)**.

@@

We evaluated the diet-associated differences in gut virome-microbiome network structure by quantifying how central each sample's network was on average. We accomplished this by utilizing two common centrality metrics: degree centrality and closeness centrality. Degree centrality, the simplest centrality metric, was defined as the number of connections each phage made with each bacterium. We supplemented measurements of degree centrality with measurements of closeness centrality. Closeness centrality is a metric of how close each phage or bacterium is to all of the other phages and bacteria in the network. A higher closeness centrality suggests that the effects of genetic information or altered abundance would be more impactful to all other microbes in the system. A network with higher average closeness centrality also indicates an overall greater degree of connections, which suggests a greater resilience against instability. Because these values are assigned to each phage and bacterium within each sample network, we calculated the average connectedness per sample and corrected for the maximum potential degree of connectedness to obtain a single value for sample graph connectedness.

@@

## Individuality of Microbial Networks
Skin and gut community membership and diversity are highly personal, with people remaining more similar to themselves than to other people over time [@Grice:2009ee; @Hannigan:2015fz; @Minot:2013ih]. We therefore hypothesized that this personal conservation extended to microbiome network structure. we addressed this hypothesis by calculating the degree of dissimilarity between each subject's network, based on phage and bacteria abundance as well as centrality. We quantified phage and bacteria centrality within each sample graph using the weighted eigenvector centrality metric. This metric defines central phages as those that are highly abundant and infect many distinct bacteria which themselves are abundant and infected by many other phages. Similarly, bacterial centrality was defined as those bacteria that were both abundant and connected to numerous phages that were themselves connected to many bacteria. We then calculated the similarity of community networks using the weighted eigenvector centrality of all nodes between all samples. Samples with similar network structures were interpreted as having similar capacities for maintaining stability and transmitting genetic material.

We used this network dissimilarity metric to test whether microbiome network structures were more similar within people than between people over time. We found that gut microbiome network structures clustered by person (ANOSIM p-value = `r signif(interstats[c(interstats$site %in% "DietAnosim"), "prob"], digits = sigfig)`, R = `r signif(interstats[c(interstats$site %in% "AS"), "prob"], digits = sigfig)`, **Figure \ref{intradiv} A**). Network dissimilarity within each person over the 8-10 day sampling period was less than the average dissimilarity between that person and others, although this difference was not statistically significant (p-value = `r signif(interstats[c(interstats$site %in% "Diet"), "prob"], digits = sigfig)`, **Figure \ref{intradiv} B**). The lack of statistical confidence was likely due to the small sample size of this dataset. Although there was evidence for gut network conservation among individuals, we found no evidence for conservation of gut network structures within families. The gut network structures were not more similar within families (twins and their mothers; intrafamily) compared to other families (inter-family) (p-value = `r signif(interstats[c(interstats$site %in% "Twins"), "prob"], digits = sigfig)`, **Figure \ref{intradiv} C**).

@@

We found evidence that diet was sufficient to alter gut microbiome network connectivity. Although our sample size was small, our findings suggested that high-fat diets were less connected than low-fat diets. This suggested that high-fat diets may lead to less stable communities with a decreased ability for microbes to directly influence one other. We supported this finding with the observation that obesity was associated with decreased network connectivity (closeness of bacterial and phage vertices). Together these findings suggest the food we eat may not only impact which microbes colonize our guts, but may also impact their interactions. Further work will be required to characterize these relationships with a larger cohort.

In addition to diet, the skin microbiome network structure varied by skin environment. Network structure differed between environmentally exposed and occluded skin sites. The sites under greater environmental fluctuation and exposure, the exposed and intermittently exposed sites, were less connected and therefore were predicted to have a higher propensity for instability. Likewise, intermittently moist sites demonstrated less connectedness than the more stable moist and sebaceous sites. Togteher this data suggested body sites under greater degrees of fluctuation harbored less connected, potentially less stable microbiomes.

Our results represent a step toward understanding the microbiome through interspecies relationships. Although these findings are informative, there are certainly caveats that should be noted. First, although our infection classification model is advantageous over existing models, we recognize that, like most classification models, ours is only as good as its training set. Large scale experimental screens for phage and bacteria infectious interactions will provide more robust model training and therefore improved model performance. Just as we have improved on previous modeling efforts, we expect that new and creative scoring metrics will be integrated into this model to further improve performance. Second, although our analyses offer an informative proof of concept, this work was done retrospectively and relied on published research from as long as seven years ago. These archived datasets were limited by the technology and costs of the time, meaning the datasets were poorly powered for the statistical analysis we strive for today. Despite these limitations, we were able to present initial conclusions; however, follow-up studies will be required to validate our findings. These methods and results will be important for informing design and interpretation of future studies and for highlighting the importance of bacteria-virus interaction profiles as a characteristic of human-associated microbial communities.

@@

# Acknowledgments
We thank the members of the Schloss lab for their underlying contributions. GDH was supported in part by the Molecular Mechanisms in Microbial Pathogenesis Training Program (T32 AI007528). GDH and PDS were supported in part by funding from the NIH (P30DK034933, U19AI09087, and U01AI124255).
